---
title: "Reproducible Research: Peer Assessment 2"
output: html_document
keep_md: true
author: Thomas Storm
date: 2024-02-12 
---
## Set knitr options

Set knitr options to not show warnings and messages to make the document less verbose.

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 
```

## Loading required packages

Data processing will be done with dplyr and lubridate, therefore loading the tidyverse package. Plot are done with lattice package.

```{r load packages}
library(tidyverse)
library(lattice)
library(R.utils)
```


## Loading and preprocessing the data

### Download data

Data are downloaded using link provided in instructions. Code checks whether data sub-directory exists, and if not, downloads and unzips file in data directory.

```{r download data}

main_dir <- getwd()
sub_dir <- "data1"

## check if "data" sub-directory exists, if not download and unzip files in data folder
if (!dir.exists(sub_dir)){
  url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  temp <- tempfile()
  dir.create(paste(main_dir, sub_dir, sep ="/"))
  download.file(url, temp)
  
  bunzip2(filename = temp, destname = file.path(paste(main_dir, sub_dir,"StormData.csv", sep ="/")), remove = FALSE)
  unlink(c(temp))
  rm(temp)
} else {print("data available")}

```

Data are downloaded and in the data folder.

### Read data

Read data with read.csv 

```{r read data, cache=TRUE}

StormData <- read.csv("./data1/StormData.csv", header = TRUE)

```

### Have a look

```{r check data}
StormData1 <- StormData[1:5000, ]

StormData1$BGN_DATE <- gsub(' 0:00:00','', StormData1$BGN_DATE)

StormData1$BGN_DATE <- mdy(StormData1$BGN_DATE)
str(StormData1)
head(StormData1)
unique(StormData1$EVTYPE)
plot(StormData1$BGN_DATE, StormData1$FATALITIES)
```

